### 2024-07-16 journal/algorithm.tex
cleaning tex files (removing comments etc)

An \opemph{Enqueue}(\var{e}) 
\rev{creates a new block at line \ref{enqNew} and sets its \fld{sum} fields to indicate
that the block represents an \op{Enqueue}.  The argument \var{e} of the \enqueue\ is stored in the block's \fld{element}
field.
The call to \op{Append}, described below, on line \ref{enqAppend}
appends the \block\ to the process's leaf and ensures it is propagated to the root.}

A \opemph{Dequeue} also 
\rev{creates a new block at line \ref{deqNew} and sets its \fld{sum} fields to indicate
that the block represents a \dequeue.
It then calls \op{Append} at line \ref{deqAppend} to append the \block\ to the process's leaf and ensure
it is propagated to the root.
Then,} the \dequeue\ calls \op{IndexDequeue} to compute
its position in the root 
 and then calls \op{FindResponse} to compute its response.

\opemph{Append}(\var{B}) first adds the block \var{B} to the invoking process's leaf.
The leaf's \fld{head} field stores the first empty slot in the leaf's \fld{blocks} array,
so the \op{Append} writes \var{B} there \rev{at line \ref{appendLeaf}} and then increments \fld{head} \rev{at line \ref{appendEnd}}.
Since \op{Append} writes only to the process's own leaf, there cannot be concurrent updates to a leaf.
\op{Append} then calls \op{Propagate} \rev{at line \ref{appendProp}} to ensure the operation represented by \var{B} is propagated to the root.

@@

A \opemph{Refresh} on node \var{v} creates a block representing the new blocks
in \var{v}'s
children and tries to append it to \var{v}.\fld{blocks}. 
Line \ref{readHead} reads \var{v}.\fld{head} into the local variable \var{h}.
\rev{Lines \ref{startHelpChild1}--\ref{endHelpChild1} ensure that the \fld{head} values
for \var{v}'s children have been advanced to the index of the latest block stored in their \fld{blocks} arrays.}
Line \ref{invokeCreateBlock} creates the new block $B$ to install in \var{v}.\fld{blocks}[\var{h}].
If line \ref{invokeCreateBlock} returns \nl\ instead of a new block, there were no new blocks in \var{v}'s children to propagate to \var{v}, so \op{Refresh} can return true at line \ref{addOP} and terminate.
Otherwise, the CAS at line \ref{cas} tries to install $B$ into \var{v}.\fld{blocks}[\var{h}].
Either this CAS succeeds or some other process has installed a  block in this location.
Either way, line \ref{advance} then calls \opemph{Advance} to advance \var{v}'s head index 
from \var{h} to $\var{h}+1$
and fill in the \fld{super} field of the most recently appended block.
The boolean value returned by \op{Refresh} indicates whether its CAS succeeded.
A \op{Refresh} may pause between a successful CAS at line \ref{cas} and calling \op{Advance} at line \ref{advance},
so other processes help keep \fld{head} up to date by  calling \op{Advance}, 
either at line \ref{helpAdvance} during a \op{Refresh} on \var{v}'s parent or line \ref{advance} during a \op{Refresh} on~\var{v}.

\opemph{CreateBlock}(\var{v, i}) is used
by \op{Refresh} to construct a new block $B$ to be installed in \var{v}.\fld{blocks}[\var{i}].
The \eleft\ and \eright\ fields store the indices of the last blocks appended to \var{v}'s
children, obtained by reading the \fld{head} index in \var{v}'s children.
Since the \fld{sum\sub{enq}} field should store the number of \enqueues\ in
\var{v}.\fld{blocks}[1..\var{i}] and these \enqueues\ come from \var{v}.\fld{left.blocks}[1..\var{B}.\eleft] and \var{v}.\fld{blocks}[1..\var{B}.\eright], line \ref{createSumEnq} sets
\fld{sum\sub{enq}} to the sum of $\var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{enq}}$ and $\var{v}.\fld{right}.\fld{blocks}[\var{B}.\eright].\fld{sum\sub{enq}}$.
Line \ref{computeNumEnq} sets \var{num\sub{enq}} to the number of \enqueues\ in the new block by
subtracting  the number of \enqueues\ in \var{v}.\fld{blocks}[$1..\var{i}-1$] from \var{B}.\fld{sum\sub{enq}}.
The values of \var{B}.\fld{sum\sub{deq}} and \var{num\sub{deq}} are computed similarly.
\rev{If the new block contains no operations, \op{CreateBlock} returns \nl\ at line \ref{testEmpty} to indicate
 there is no need to install~it.
Otherwise,} if \var{B}
is going to be installed in the root, line \ref{computeLength} \rev{uses Equation (\ref{size-eqn}) to compute $B$'s} \fld{size} field, which
represents the number of elements in the queue after the operations in the block are performed.

>>>>

An \opemph{Enqueue}(\var{e}) 
creates a new block at line \ref{enqNew} and sets its \fld{sum} fields to indicate
that the block represents an \op{Enqueue}.  The argument \var{e} of the \enqueue\ is stored in the block's \fld{element}
field.
The call to \op{Append}, described below, on line \ref{enqAppend}
appends the \block\ to the process's leaf and ensures it is propagated to the root.

A \opemph{Dequeue} also 
creates a new block at line \ref{deqNew} and sets its \fld{sum} fields to indicate
that the block represents a \dequeue.
It then calls \op{Append} at line \ref{deqAppend} to append the \block\ to the process's leaf and ensure
it is propagated to the root.
Then, the \dequeue\ calls \op{IndexDequeue} to compute
its position in the root 
 and then calls \op{FindResponse} to compute its response.

\opemph{Append}(\var{B}) first adds the block \var{B} to the invoking process's leaf.
The leaf's \fld{head} field stores the first empty slot in the leaf's \fld{blocks} array,
so the \op{Append} writes \var{B} there at line \ref{appendLeaf} and then increments \fld{head} at line \ref{appendEnd}.
Since \op{Append} writes only to the process's own leaf, there cannot be concurrent updates to a leaf.
\op{Append} then calls \op{Propagate} at line \ref{appendProp} to ensure the operation represented by \var{B} is propagated to the root.

@@

A \opemph{Refresh} on node \var{v} creates a block representing the new blocks
in \var{v}'s
children and tries to append it to \var{v}.\fld{blocks}. 
Line \ref{readHead} reads \var{v}.\fld{head} into the local variable \var{h}.
Lines \ref{startHelpChild1}--\ref{endHelpChild1} ensure that the \fld{head} values
for \var{v}'s children have been advanced to the index of the latest block stored in their \fld{blocks} arrays.
Line \ref{invokeCreateBlock} creates the new block $B$ to install in \var{v}.\fld{blocks}[\var{h}].
If line \ref{invokeCreateBlock} returns \nl\ instead of a new block, there were no new blocks in \var{v}'s children to propagate to \var{v}, so \op{Refresh} can return true at line \ref{addOP} and terminate.
Otherwise, the CAS at line \ref{cas} tries to install $B$ into \var{v}.\fld{blocks}[\var{h}].
Either this CAS succeeds or some other process has installed a  block in this location.
Either way, line \ref{advance} then calls \opemph{Advance} to advance \var{v}'s head index 
from \var{h} to $\var{h}+1$
and fill in the \fld{super} field of the most recently appended block.
The boolean value returned by \op{Refresh} indicates whether its CAS succeeded.
A \op{Refresh} may pause between a successful CAS at line \ref{cas} and calling \op{Advance} at line \ref{advance},
so other processes help keep \fld{head} up to date by  calling \op{Advance}, 
either at line \ref{helpAdvance} during a \op{Refresh} on \var{v}'s parent or line \ref{advance} during a \op{Refresh} on~\var{v}.

\opemph{CreateBlock}(\var{v, i}) is used
by \op{Refresh} to construct a new block $B$ to be installed in \var{v}.\fld{blocks}[\var{i}].
The \eleft\ and \eright\ fields store the indices of the last blocks appended to \var{v}'s
children, obtained by reading the \fld{head} index in \var{v}'s children.
Since the \fld{sum\sub{enq}} field should store the number of \enqueues\ in
\var{v}.\fld{blocks}[1..\var{i}] and these \enqueues\ come from \var{v}.\fld{left.blocks}[1..\var{B}.\eleft] and \var{v}.\fld{blocks}[1..\var{B}.\eright], line \ref{createSumEnq} sets
\fld{sum\sub{enq}} to the sum of $\var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{enq}}$ and $\var{v}.\fld{right}.\fld{blocks}[\var{B}.\eright].\fld{sum\sub{enq}}$.
Line \ref{computeNumEnq} sets \var{num\sub{enq}} to the number of \enqueues\ in the new block by
subtracting  the number of \enqueues\ in \var{v}.\fld{blocks}[$1..\var{i}-1$] from \var{B}.\fld{sum\sub{enq}}.
The values of \var{B}.\fld{sum\sub{deq}} and \var{num\sub{deq}} are computed similarly.
If the new block contains no operations, \op{CreateBlock} returns \nl\ at line \ref{testEmpty} to indicate
 there is no need to install~it.
Otherwise, if \var{B}
is going to be installed in the root, line \ref{computeLength} uses Equation (\ref{size-eqn}) to compute $B$'s \fld{size} field, which
represents the number of elements in the queue after the operations in the block are performed.

### 2024-07-16 journal/analysis.tex
cleaning tex files (removing comments etc)

\begin{mytheorem}\label{enqDeqTime}
\rev{The queue implementation is wait-free and}
each \op{Enqueue} and null Dequeue takes $O(\log p)$ steps 
and each non-null \op{Dequeue} takes
$O(\log p\log c + \log q_e+ \log q_d)$ steps,
where $q_d$ is the size of the queue when the \op{Dequeue} is linearized and 
$q_e$ is the size of the queue when the \op{Enqueue} of the value returned is linearized.
\end{mytheorem}
\begin{proof}
An \op{Enqueue} or null \op{Dequeue} creates a block, appends it to the process's 
leaf and propagates it to the root.  The \op{Propagate}  does $O(1)$ steps 
at each node on the path from the process's leaf to the root.
A null \op{Dequeue} additionally calls \op{IndexDequeue}, which also does $O(1)$ steps
at each node on this path. 
So, the total number of steps for either type of operation is $O(\log p)$.

>>>>

\begin{theorem}\label{enqDeqTime}
The queue implementation is wait-free and
each \op{Enqueue} and null Dequeue takes $O(\log p)$ steps 
and each non-null \op{Dequeue} takes
$O(\log p\log c + \log q_e+ \log q_d)$ steps,
where $q_d$ is the size of the queue when the \op{Dequeue} is linearized and 
$q_e$ is the size of the queue when the \op{Enqueue} of the value returned is linearized.
\end{theorem}
\begin{proof}
An \op{Enqueue} or null \op{Dequeue} creates a block, appends it to the process's 
leaf and propagates it to the root.  The \op{Propagate}  does $O(1)$ steps 
at each node on the path from the process's leaf to the root.
A null \op{Dequeue} additionally calls \op{IndexDequeue}, which also does $O(1)$ steps
at each node on this path. 
So, the total number of steps for either type of operation is $O(\log p)$.

### 2024-07-16 journal/conclusion.tex
cleaning tex files (removing comments etc)

Building on the work described in this paper, Asbell and Ruppert \cite{AR23} have
designed a doubly-ended queue with polylogarithmic amortized step
complexity, which also yields a stack as a special case.  
This required a substantially different representation of the data stored in the ordering tree.
Whether the ordering tree could also be used to obtain a priority queue with polylogarithmic step complexity
remains an open question.
\rev{The technique of cooperatively propagating information up a tree with a double refresh
at each node was also used recently to obtain lock-free augmented search trees \cite{FR24}.}

>>>>

Building on the work described in this paper, Asbell and Ruppert \cite{AR23} have
designed a doubly-ended queue with polylogarithmic amortized step
complexity, which also yields a stack as a special case.  
This required a substantially different representation of the data stored in the ordering tree.
Whether the ordering tree could also be used to obtain a priority queue with polylogarithmic step complexity
remains an open question.
The technique of cooperatively propagating information up a tree with a double refresh
at each node was also used recently to obtain lock-free augmented search trees \cite{FR24}.

### 2024-07-16 journal/correctness-GC.tex
cleaning tex files (removing comments etc)

If line \ref{casGC} of a \op{Refresh} modifies $\var{v}.\fld{blocks}$, then \var{v} is an internal node.
After reading $T$ from $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC},
the \op{Refresh} then creates the block \var{B},
and calls \op{AddBlock}($v,T,B$) to create $T'$.
Line \ref{createIndexGC} sets $\var{B}.\var{index} = \max(I)+1$.
The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT \rev{at line \ref{splitGC}}
and then add $B$ to it \rev{at line \ref{GCend} or \ref{noGC-insert}}.
So the claim is satisfied.
\end{proof}

@@

%\here{If next obs is only used once, can fold it into the proof that needs it}
%\begin{observation}
%\label{edirGC} 
%If $B$ is a block in the RBT of an internal node \var{v}, then for $\var{dir} \in \{\fld{left},\fld{right}\}$,
%$B.\edir$ was read from the \fld{index} field of some node in the RBT of $\var{v}.\var{dir}$.
%\end{observation}
%\begin{proof}
%The \eleft\ and \eright\ fields are set by reading the \fld{index} field of a node in the appropriate child
%on line \ref{createEndLeftGC} and \ref{createEndRightGC}.
%\end{proof}

@@

\begin{customlemma}{\ref{lem::doubleRefresh}$'$}\label{lem::doubleRefreshGC}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}(\var{v}) by the same process.
All operations contained \var{v}'s children when $R_1$ begins
are contained in \var{v} when $R_2$ terminates.
\end{customlemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{casGC} or terminates at line \ref{addOPGC}, the claim follows
from Lemma \ref{successfulRefreshGC}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{casGC}.
Then some other \op{CAS} on $\var{v}.\fld{blocks}$ succeeds between \rev{each \op{Refresh}'s 
read of $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC} and its \op{CAS}} at line \ref{casGC}.
Consider the \op{Refresh} $R_3$ that does this successful \op{CAS} during $R_2$.
$R_3$ must have read $\var{v}.\fld{blocks}$ after the successful \op{CAS} during $R_1$.
The claim follows from Lemma \ref{successfulRefreshGC} applied to~$R_3$.
\end{proof}

@@

%\here{Technically, following proof might have to be wrapped up in a big induction with the argument
%that outputs are linearized correctly, to argue that the matchup between \enqueues\ and \dequeues\ 
%is the same as it would be in the linearization ordering.  (Check whether this is really needed.)
%Or alternatively, we could add an assumption that if every operation agrees with the linearization
%so far, then the invariant is preserved by each step.
%Then, when proving the linearization is correct, everything will work out.
%But just leave it as is for the submission, since this is a subtle point}
% NO, I DON'T THINK THIS IS NECESSARY

@@

\begin{lemma}\label{blocknotfound}
If a \op{Dequeue} operation fails to find a block in an RBT,
%during its calls to \op{IndexDequeue} or \op{GetEnqueue}, 
then \rev{the \dequeue} has been propagated to the root and its result has been written in the \fld{response} field of the leaf block it created.
If an \op{Enqueue} operation fails to find a block in an RBT, it has been propagated to the root.
\end{lemma}
\begin{proof}
Any block that GC removes from an RBT is finished, by Invariant \ref{inv::finished}, so if an 
\op{Enqueue} or \op{Dequeue} fails to find a block while it is propagating itself up to
the root (for example, during the \op{CreateBlock} routine), then a block containing
the operation itself has been removed from a RBT, so the operation has propagated to the root,
by the definition of finished.
Moreover, by Invariant \ref{inv::finished}, if the operation is a \dequeue, then its result is in its \fld{response} field.

@@

We first bound the size of RBTs.  Let $q_{max}$ be the maximum size of the queue at any time during the sequential execution given by the linearization $L$.
Recall that GC is done on a node every $G$ times its RBT is updated, and we chose $G$ to be $p^2\ceil{\log p}$.
Part of the proof of the following lemma is similar to the proof of Lemma \ref{dSearchTime}.
%We say the index $b$ \emph{is marked in node \var{v}} if some process has tested
%whether $b>\var{v}.\fld{last}[*]$ at line \ref{marktest1} or \ref{marktest2}.
%Intuitively, a block's index value is marked 
%when a \dequeue\ passes through this block to find the \enqueue
%whose argument it should return, or when a \dequeue\ that returns null passes through the block.
%Thus, $b$ gets marked in \var{v} when $\var{v}.\fld{blocks}[b]$ 
%contains either an \enqueue whose value is being dequeued or a \nl\ \dequeue.
%If the index of block $B$ is marked in \var{v}, we also say $B$ is marked in \var{v}.

\begin{lemma}\label{boundAfterGC}
If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
it contains at most $3q_{max}+5p+1$ blocks.
\end{lemma}
\begin{proof}
Consider the invocation $A$ of \op{AddBlock} that updates a node \var{v}'s RBT with the insertion of a block whose 
\fld{index} is a multiple of $G$.
Then, $A$ performs a GC phase.
Let \rev{$s$ be the step} when $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
That call to \op{SplitBlock} recurses up to the root, where it computes $m$ by reading the \var{last}
array.
Let $L_1$ be the prefix of the linearization $L$ corresponding to blocks $1..m$ of the root.
Let $L_2$ be the next segment of the linearization corresponding to blocks $m+1..\ell$ of the root, where
$\ell$ is the last block added to the root's RBT before $s$.

@@

The following theorem bounds the space that is reachable (and therefore cannot be freed by the environment's garbage collector) at any time.
%\here{At first, I thought the following analysis could be tightened
%by using the fact that operations only appear along one path in tree
%so doing a bound on total size of all nodes on one level might work yield better bound.
%But I don't think so, because we could have each process go in order to dequeue
%everything and then enqueue q new items (where q>>p).  Then sets of enqueue operations
%stored in each node at one level of tree would be disjoint (because of the way
%GC is timed differently at each node).

\begin{mytheorem}\label{spaceBound}
The queue data structure uses a maximum of $O(pq_{max}+p^3\log p)$ words of memory at any time.
\end{mytheorem}
\begin{proof}
There are $2p-1$ nodes in the ordering tree.  Aside from the RBT, each node uses $O(1)$ memory words.
Each process may hold pointers to $O(1)$ RBTs that are no longer current in local variables.
So the space bound follows from Corollary \ref{RBTbound} and the fact that $G$ is chosen
to be $p^2\ceil{\log p}$.
\end{proof}

@@

\begin{mytheorem}
The amortized step complexity of each operation is $O(\log p \log(p+q_{max}))$.
\end{mytheorem}
\begin{proof}
It follows from Corollary \ref{RBTbound} and our choice of $G=p^2\ceil{\log p}$
that all the RBT routines we use to perform \op{Split}, \op{Insert} and searches for
blocks with a particular index or for a \fld{sum\sub{enq}} value (in line \ref{FRsearchGC} or \ref{getChildGC}) can  be done in $O(\log(p+q_{max}))$ steps.

>>>>

If line \ref{casGC} of a \op{Refresh} modifies $\var{v}.\fld{blocks}$, then \var{v} is an internal node.
After reading $T$ from $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC},
the \op{Refresh} then creates the block \var{B},
and calls \op{AddBlock}($v,T,B$) to create $T'$.
Line \ref{createIndexGC} sets $\var{B}.\var{index} = \max(I)+1$.
The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT at line \ref{splitGC}
and then add $B$ to it at line \ref{GCend} or \ref{noGC-insert}.
So the claim is satisfied.
\end{proof}

@@

@@

\begin{customlemma}{\ref{lem::doubleRefresh}$'$}\label{lem::doubleRefreshGC}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}(\var{v}) by the same process.
All operations contained \var{v}'s children when $R_1$ begins
are contained in \var{v} when $R_2$ terminates.
\end{customlemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{casGC} or terminates at line \ref{addOPGC}, the claim follows
from Lemma \ref{successfulRefreshGC}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{casGC}.
Then some other \op{CAS} on $\var{v}.\fld{blocks}$ succeeds between each \op{Refresh}'s 
read of $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC} and its \op{CAS} at line \ref{casGC}.
Consider the \op{Refresh} $R_3$ that does this successful \op{CAS} during $R_2$.
$R_3$ must have read $\var{v}.\fld{blocks}$ after the successful \op{CAS} during $R_1$.
The claim follows from Lemma \ref{successfulRefreshGC} applied to~$R_3$.
\end{proof}

@@

@@

\begin{lemma}\label{blocknotfound}
If a \op{Dequeue} operation fails to find a block in an RBT,
%during its calls to \op{IndexDequeue} or \op{GetEnqueue}, 
then the \dequeue\ has been propagated to the root and its result has been written in the \fld{response} field of the leaf block it created.
If an \op{Enqueue} operation fails to find a block in an RBT, it has been propagated to the root.
\end{lemma}
\begin{proof}
Any block that GC removes from an RBT is finished, by Invariant \ref{inv::finished}, so if an 
\op{Enqueue} or \op{Dequeue} fails to find a block while it is propagating itself up to
the root (for example, during the \op{CreateBlock} routine), then a block containing
the operation itself has been removed from a RBT, so the operation has propagated to the root,
by the definition of finished.
Moreover, by Invariant \ref{inv::finished}, if the operation is a \dequeue, then its result is in its \fld{response} field.

@@

We first bound the size of RBTs.  Let $q_{max}$ be the maximum size of the queue at any time during the sequential execution given by the linearization $L$.
Recall that GC is done on a node every $G$ times its RBT is updated, and we chose $G$ to be $p^2\ceil{\log p}$.
Part of the proof of the following lemma is similar to the proof of Lemma \ref{dSearchTime}.

\begin{lemma}\label{boundAfterGC}
If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
it contains at most $3q_{max}+5p+1$ blocks.
\end{lemma}
\begin{proof}
Consider the invocation $A$ of \op{AddBlock} that updates a node \var{v}'s RBT with the insertion of a block whose 
\fld{index} is a multiple of $G$.
Then, $A$ performs a GC phase.
Let $s$ be the step when $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
That call to \op{SplitBlock} recurses up to the root, where it computes $m$ by reading the \var{last}
array.
Let $L_1$ be the prefix of the linearization $L$ corresponding to blocks $1..m$ of the root.
Let $L_2$ be the next segment of the linearization corresponding to blocks $m+1..\ell$ of the root, where
$\ell$ is the last block added to the root's RBT before $s$.

@@

The following theorem bounds the space that is reachable (and therefore cannot be freed by the environment's garbage collector) at any time.

\begin{theorem}\label{spaceBound}
The queue data structure uses a maximum of $O(pq_{max}+p^3\log p)$ words of memory at any time.
\end{theorem}
\begin{proof}
There are $2p-1$ nodes in the ordering tree.  Aside from the RBT, each node uses $O(1)$ memory words.
Each process may hold pointers to $O(1)$ RBTs that are no longer current in local variables.
So the space bound follows from Corollary \ref{RBTbound} and the fact that $G$ is chosen
to be $p^2\ceil{\log p}$.
\end{proof}

@@

\begin{theorem}
The amortized step complexity of each operation is $O(\log p \log(p+q_{max}))$.
\end{theorem}
\begin{proof}
It follows from Corollary \ref{RBTbound} and our choice of $G=p^2\ceil{\log p}$
that all the RBT routines we use to perform \op{Split}, \op{Insert} and searches for
blocks with a particular index or for a \fld{sum\sub{enq}} value (in line \ref{FRsearchGC} or \ref{getChildGC}) can  be done in $O(\log(p+q_{max}))$ steps.

### 2024-07-16 journal/related.tex
cleaning tex files (removing comments etc)

\paragraph{Universal Constructions.}
\Eric{Rewrote this paragraph on Oct 25}
One can also build a queue using a universal construction \cite{10.1145/114005.102808}.
Afek, Dauber, and Touitou's universal construction~\cite{DBLP:conf/stoc/AfekDT95}
introduced the technique where a process must climb a binary tree while helping any other processes it sees along the way to reach the root,
and this technique is the basis of our queue implementation.
Jayanti \cite{DBLP:conf/podc/Jayanti98a} observed that
their construction can be modified to use $O(\log p)$ steps per operation, 
assuming that words can store $\Omega(p \log p)$ bits. 
However, if more reasonably-sized $O(\log p)$-bit words are used, the construction would take $\Omega(p\log p)$ steps per operation.
There are two obstacles to making this construction more efficient:
it is expensive to manipulate lists of up to $p$ operations that must be propagated along a path
up to the root, and
when a batch of up to $p$ operations reaches the root, it is expensive to perform all of them on the
implemented data structure.
In our work, we devise a novel way of doing this implicitly for batches of operations on a queue.

@@

\paragraph{Other Primitives.}
Li \cite{Li01} implemented a non-blocking linearizable queue using only the weak primitives
test\&set, fetch\&add and swap.
This implementation's amortized step complexity is not bounded
as a function of the number of processes and the size of the queue.
(It depends on the total number of enqueues in the execution.)

\HERE{discuss \cite{ACH18} who seem to have showed that strongly linearizable (define) queue is impossible using weak primitives; can also mention that HW22 observed that the MS queue is not strongly linearizable but can easily be made so}

>>>>

\paragraph{Universal Constructions.}
One can also build a queue using a universal construction \cite{10.1145/114005.102808}.
Afek, Dauber, and Touitou's universal construction~\cite{DBLP:conf/stoc/AfekDT95}
introduced the technique where a process must climb a binary tree while helping any other processes it sees along the way to reach the root,
and this technique is the basis of our queue implementation.
Jayanti \cite{DBLP:conf/podc/Jayanti98a} observed that
their construction can be modified to use $O(\log p)$ steps per operation, 
assuming that words can store $\Omega(p \log p)$ bits. 
However, if more reasonably-sized $O(\log p)$-bit words are used, the construction would take $\Omega(p\log p)$ steps per operation.
There are two obstacles to making this construction more efficient:
it is expensive to manipulate lists of up to $p$ operations that must be propagated along a path
up to the root, and
when a batch of up to $p$ operations reaches the root, it is expensive to perform all of them on the
implemented data structure.
In our work, we devise a novel way of doing this implicitly for batches of operations on a queue.

@@

\paragraph{Queues Implemented from Other Primitives.}
Li \cite{Li01} implemented a non-blocking linearizable queue using only the weak primitives
test\&set, fetch\&add and swap.
This implementation's amortized step complexity is not bounded
as a function of the number of processes and the size of the queue.
(It depends on the total number of enqueues in the execution.)

%Could discuss \cite{ACH18} who seem to have showed that strongly linearizable 
%(define) queue is impossible using weak primitives; can also mention that HW22 
%observed that the MS queue is not strongly linearizable but can easily be made so}
